Side effects

### 2020-08-27 inst/analysis/side-effects.Rmd
Percentages for AST size

options(repos="https://cran.r-project.org")
theme_set(theme_minimal())
create_tags(path(TAGS_DIR, "side-effects.tex"), prefix="", default=TRUE)
```

```{r}
core_data <- read_fst(EVALS_SUM_CORE_FILE)
package_data <- read_fst(EVALS_SUM_PKGS_FILE)
kaggle_data <- read_fst(EVALS_SUM_KAGGLE_FILE)
program_data <- read_fst(PROGRAM_FILE)

all_data <- rbind(core_data, package_data, kaggle_data)
```

## Side-Effects and Eval

Each entry in `call.fst` contains `direct_writes` and `indirect_writes` (both
`0` for `eval.parent` and `local`). `direct_writes` are the side-effects
directly caused by the `eval` in question and `indirect_writes` are the
side-effects caused by all children evals of the eval. All application
assignments can be found in `direct_writes` field of `program.fst` and all
`eval_assignments` can be found in `indirect_writes` field of `program.fst`. All
non-local assignments can be found by simply summing up the `direct_writes`
field of all eval calls.

```{r preprocess-write-data}
eval_writes_data <-
    all_data %>%
    filter(eval_function == "eval" | eval_function == "evalq") %>%
    select(eval_source_type, eval_source, eval_call_srcref, direct_writes, indirect_writes, nb_ev_calls) %>%
    mutate(all_writes = direct_writes + indirect_writes, side_effecting = as.logical(all_writes))

program_writes_data <-
    program_data %>%
    select(direct_writes, indirect_writes)
```

```{r compute-side-effecting-write-counts}
all_writes <- sum(program_writes_data$direct_writes)
eval_writes <- sum(program_writes_data$indirect_writes)

r("All Writes", all_writes)
r("Eval Writes", eval_writes)

eval_side_effecting_writes <- sum(eval_writes_data$direct_writes * eval_writes_data$nb_ev_calls)

r("Eval Side Effecting Writes", eval_side_effecting_writes)

r("Eval Side Effecting Writes Eval Perc", ratio(eval_side_effecting_writes, eval_writes))
r("Eval Side Effecting Writes All Perc", ratio(eval_side_effecting_writes, all_writes))
```

```{r compute-side-effecting-call-counts}
eval_calls_by_source_writes <-
    eval_writes_data %>%
    count(eval_source_type, side_effecting, wt = nb_ev_calls, name = "call_count") %>%
    group_by(eval_source_type) %>%
    mutate(total_call_count = sum(call_count)) %>%
    ungroup() %>%
    mutate(call_perc = round(100 * call_count/total_call_count, 1)) %>%
    filter(side_effecting)

@@

eval_calls_by_source_writes %>%
pmap(function(eval_source_type, side_effecting, call_count, total_call_count, call_perc) {
    r(str_c("Side Effecting ", eval_source_type, " call perc"), str_c(call_perc, "%"))
})

eval_calls_by_package_writes <-
    eval_writes_data %>%
    filter(eval_source_type != "core") %>%
    count(eval_source, side_effecting, wt = nb_ev_calls, name = "call_count") %>%
    arrange(desc(call_count))

datatable(eval_calls_by_package_writes)
```

The call site computation is incorrect.

```{r compute-side-effecting-site-counts, eval = FALSE}
eval_sites_by_source_writes <-
    eval_writes_data %>%
    group_by(eval_source_type, eval_call_srcref) %>%
    summarize(side_effecting = any(side_effecting)) %>%
    ungroup() %>%
    count(eval_source_type, side_effecting, name = "site_count") %>%
    group_by(eval_source_type) %>%
    mutate(total_site_count = sum(site_count)) %>%
    ungroup() %>%
    mutate(site_perc = round(100 * site_count / total_site_count, 1)) %>%
    filter(side_effecting)

@@

eval_sites_by_source_writes %>%
pmap(function(eval_source_type, side_effecting, site_count, total_site_count, site_perc) {
    r(str_c("SideEffecting ", eval_source_type, " site perc"), str_c(site_perc, "%"))
})

eval_sites_by_package_writes <-
    eval_writes_data %>%
    filter(eval_source_type != "core") %>%
    group_by(eval_source, eval_call_srcref) %>%
    summarize(side_effecting = any(side_effecting)) %>%
    ungroup() %>%
    count(eval_source, side_effecting, name = "site_count") %>%
    group_by(eval_source, side_effecting) %>%
    mutate(site_perc = round(100 * site_count / sum(site_count), 1)) %>%
    ungroup() %>%
    arrange(side_effecting, desc(site_perc))

datatable(eval_sites_by_package_writes)
```

>>>>

options(repos="https://cran.r-project.org")
theme_set(theme_minimal())
#create_tags(path(TAGS_DIR, "side-effects.tex"), prefix="SE", default=TRUE)
```

```{r}
# FIXME: path!
calls_raw <- read_fst(print(path(params$base_dir, "run", "package-evals-traced.1", "calls.fst")))
calls <- 
  calls_raw %>%
  mutate(package=basename(dirname(dirname(file))))

core_data <- read_fst(EVALS_SUM_CORE_FILE)
package_data <- read_fst(EVALS_SUM_PKGS_FILE)
kaggle_data <- read_fst(EVALS_SUM_KAGGLE_FILE)
program_data <- read_fst(PROGRAM_FILE)

all_data <- rbind(core_data,
                  package_data, 
                  kaggle_data)

head(all_data)

```

## Amount of computation via eval

```{r writes-data}
writes_data <-
    all_data %>%
    select(eval_source_type, eval_source, eval_call_srcref, direct_writes, indirect_writes) %>%
    filter(eval_function == "eval" | eval_function == "evalq") %>%
    mutate(all_writes = direct_writes + indirect_writes, side_effecting = as.logical(all_writes))
# eval_events <-
#  calls %>%
#  filter(!is.na(eval_call_srcref)) %>%
#  filter(eval_function == "eval" | eval_function == "evalq") %>%
#  mutate(all_writes = direct_writes + indirect_writes) %>%
#  select(package, file, eval_call_srcref, eval_call_id, builtin, special, closure, interpreter_eval, c_call, allocation, direct_writes, indirect_writes, all_writes) %>%
#  pivot_longer(cols=c(builtin, special, closure, interpreter_eval, c_call, allocation, direct_writes, indirect_writes, all_writes), names_to="event", values_to="event_count")
```

```{r side-effects-analysis}
eval_calls_by_source_writes <-
    writes_data %>%
    count(eval_source_type, side_effecting, wt = nb_ev_calls, name = call_count)

@@

eval_calls_by_package_writes <-
    writes_data %>%
    filter(eval_source_type != "core") %>%
    count(eval_source, side_effecting, wt = nb_ev_calls, name = call_count) %>%
    arrange(desc(call_count))

datatable(eval_calls_by_package_writes)

eval_sites_by_source_writes <-
    writes_data %>%
    group_by(eval_source_type, eval_call_srcref) %>%
    summarize(side_effecting = any(side_effecting)) %>%
    ungroup() %>%
    count(eval_source_type, side_effecting, name = site_count)

@@

eval_sites_by_package_writes <-
    writes_data %>%
    filter(eval_source_type != "core") %>%
    group_by(eval_source, eval_call_srcref) %>%
    summarize(side_effecting = any(side_effecting)) %>%
    ungroup() %>%
    count(eval_source, side_effecting, name = site_count) %>%
    group_by(eval_source, side_effecting) %>%
    mutate(site_perc = round(100 * site_count / sum(site_count), 1)) %>%
    ungroup() %>%
    arrange(side_effecting, desc(site_perc))

datatable(eval_sites_by_package_writes)

```

Each entry in `call.fst` contains `direct_writes` and `indirect_writes` (both `0` for `eval.parent` and `local`). 
`direct_writes` are the side-effects directly caused by this `eval`. 
`indirect_writes` are the side-effects caused by all children evals of this eval.

For graphing, you may want to sum them up for each eval.
I compute them separately because it is more informative (should we need this information for debugging or qualitative insights).
Now, for normalization you have two quantities: all assignments done by the application and all assignments done by all evals (non-local or local).
all application assignments can be found in `direct_writes` field of `program.fst` and all `eval_assignments` can be found in `indirect_writes` field of `program.fst`.
You can also find all non-local assignments by simply summing up the `direct_writes` field of all eval calls.

```{r}

```
